Modeling memory and taxis to analyze migrations

### 2021-06-02 text/ms.Rmd
Working on a new model

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: console
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996]. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].

@@

Here, we explore the relationship between memory and resource following for a migratory population under various static and dynamic seasonal resource distributions using a diffusion-advection model.

@@

We anticipate that under many conditions a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal resource environments.

@@

In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, (5) some social aspect to the learned behavior.

A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system.  We consider a population moving in one dimension in a constrained domain ${\cal D}$ and distributing itself according to the following equation:

@@

The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e. $\widehat{u}(t-\tau)$, where $\hat{\,}$ operator signifies the expected value of the population distribution. This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:

$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$

The long-term, reference memory term, which does not change throughout the evolution of the process, can be thought of a strictly periodic function.  The parameter $\kappa \in (0,1)$ is the relative contribution of that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.  Importantly, the individuals remember not just the location of the actionable memory term, but the movement rate of that term as well, i.e. $d\widehat{u_m}(t)/dt$.

The overarching behavior of the population is to attempt to repeat that memory.  Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e. for those animals that are lagging behind the optimal remembered location, the memory speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.

@@

$$v_m(t) =  \begin{cases}
\frac{\lambda}{1 + \exp((x - x_+)/\beta)}  &  \frac{d\widehat{u_m}}{dt} > 0\\
\frac{\lambda}{1 - \exp(-(x - x_-)/\beta)} & \frac{d\widehat{u_m}}{dt} < 0
\end{cases}$$

where $x_+$ and $x_-$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.

@@

The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.  Thus the total population is fixed at all times - there are no birth or death processes - and found within the domain.   Furthermore, the parameters remain constant throughout time.  Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic and changing environment.

<!--
A few special cases of this model have been studied in the past. @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.  @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process. Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model. @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function. @Fagan2019 added a behavioral switching behavior to the non-local information model.  This study 
-->

@@

We ran our numerical experiments on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).  We were interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).  Both are characterized by a peak in time and space centered at $M_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.  The island resource is simply two uncorrelated bivariate normal distributions $$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
where the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e. ${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.

The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1.  This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).

Within a given year, the resource is entirely symmetric: $h(x,t) = h(-x, \tau-t)$.  However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.  Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.  These trends mirror the poleward shift of peak resources and the earlier spring phenology occurring with a warming global climate.  In the scenarios we explore, the spatial scales ($s_x$ and $s_t$) remain constant.

@@

For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.

We also record the actual memory.

@@

We numerically solved this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.

@@

Q1. What

>>>>

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: inline
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1988]. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2014].

@@

@@

Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.  In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from a non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon (**refs**).  Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing patterns in resource distribution that are the consequences of rapid climate change.

We anticipate that under many conditions a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal environments.  Over-reliance on either strategy should be maladaptive.  We further anticipate that a a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are increasingly stochastic.

@@

In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, (5) some social aspect to the learned behavior.  
A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system.  We consider a population moving in one dimension in a constrained domain ${\cal D}$ and distributing itself according to the following equation:

@@

The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e. $\widehat{u}(t-\tau)$. This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:

$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)$$

The long-term, reference memory term, which does not change throughout the evolution of the process is a strictly periodic function: $\widehat{u_o}(t)$ = $\widehat{u_o}(t - \tau)$.  The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.  Importantly, the individuals remember not just the location of the actionable memory term, but the movement rate of that term as well, i.e. $d\widehat{u_m}(t)/dt$.

The overarching behavior of the population is to attempt to echo that social memory.  Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e. for those animals that are lagging behind the optimal remembered location, the memory speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.

@@

$$v_m(x,t) =  \begin{cases}
\frac{\lambda}{1 + \exp((x - x_+(t))/\beta)}  &\text{if} &  \frac{d\widehat{u_m}}{dt} > 0\\
\frac{\lambda}{1 - \exp(-(x - x_-(t))/\beta)} &\text{if} & \frac{d\widehat{u_m}}{dt} < 0
\end{cases}$$

where $x_+(t)$ and $x_-(t)$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.

@@

The parameter $\beta$, the reciprocal of the maximum slope of the logistic regression, can be considered a spatial scale of social cohesion.  If it is large, the population responds more or less similarly across the domain, with the extreme case $\beta = \infty$ corresponding to a shared velocity of $\frac{d\widehat{u_m}}{dt}$.  If it is small, then there is a very strong discrepancy in speeds at a short scale (Figure 1). The parameter $\lambda$ - corresponding to the maximum possible advective speed - is needed to constrain the speed at large distances from $u_m$ for numerical stability, but is not of central interest in our study.

The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.  As there are no birth or death processes, to total population remains fixed and constant, for convenience integrating to 1.  Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process.  Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' an intra- and interannually dynamic, seasonal environment.

<!-- A few special cases of this model have been studied in the past. @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.  @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process. Importantly, the authors prove the well-posedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model. @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function. @Fagan2019 added a behavioral switching behavior to the non-local information model.  The essential innovation of our study is the inclusion of an explicit, slowly-varying migratory advection term.
-->

```{r, eval = FALSE}
require(memorymigration)

x <- -100:100
m.hat <- 40
dm <- 5
beta <- 20
lambda <- 40

v1 <- getMemoryVelocity(x, m.hat, dm, beta = 20, lambda = 40)
v2 <- getMemoryVelocity(x, m.hat, dm, beta = 100, lambda = 40)
v3 <- getMemoryVelocity(x, m.hat, dm, beta = 5, lambda = 40)

png("figures/MemoryVelocity.png", width = 1600, height = 1000, res = 300)
par(mar = c(3,3,1,1), tck = 0.01, mgp = c(1.5,.25,0), bty = "l", 
    cex.lab = 1.25)
plot(x, v1, typ = "l", lwd = 2, ylab= "memory velocity v(t)", 
     col = "darkred", xlab = "spatial domain (x)", ylim = c(0,45))
lines(x, v2, col = "darkgreen", lwd = 2)
lines(x, v3, col = "darkblue", lwd = 2)
segments(-100,dm,100,dm, col = "darkgrey", lty = 3, lwd = 2)
abline(v = m.hat, col = "darkgrey", lty = 3, lwd = 2)
points(m.hat, dm, bg = "darkgrey", pch= 21, cex = 1.5)
legend("topright", col = c("darkblue", "darkred", "darkgreen"),
       title = expression(beta), lwd = 2, legend = c(5,20,100))
text(-85, dm, expression(d~widehat(u[m])/dt), pos = 3, col = grey(.3))
text(m.hat, lambda, expression(widehat(u[m])(t)), pos = 2, col = grey(.3), xpd = NA)
dev.off()
```
![](figures/MemoryVelocity.png)
**Figure 1.** *Illustration of memory-based velocity term as a logistic function of space (x-axis), mean memory ($\widehat{u_m}(t)$ - vertical dashed line) and mean movement rate ($d\widehat{u_m}/dt$ - horizontal dashed line). Various values of the social spatial scale parameter $\beta$ reflects the maximum slope of the logistic curve such that at small values of the spatial scale, there is a steep difference in the speed around the mean location (blue curve), and at large values of the spatial scale, there as a weak response of the speed to the mean location.  In all cases, at locations beyond the mean, the speed is lower than at locations lagging behind the mean.*

@@

We ran our numerical experiments on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).  We were interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).  Both are characterized by a peak in time and space centered at $M_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.  The island resource is simply two uncorrelated bivariate normal distributions $$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e. ${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.

The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1 ($\int_X h(x,t) dx = 1$).  This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).

Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.  However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.  Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.  These trends mirror the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.  The spatial scales ($s_x$ and $s_t$) remain constant.

@@

For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.  We also record the migration memory $\widehat{u_m}$ for each year.

@@

We numerically solve this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.  The complete code is available as an R package (`memorymigration`) available on GitHub at https://github.com/EliGurarie/memorymigration and as an interactive Shiny application at https://url.to.be.determined.

@@
